---
title: "assingment"
author: "Catherine Rohner"
date: "2024-03-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressPackageStartupMessages({
  library(AnnotationHub)
  library(rtracklayer)
  library(epiwraps)
  library(GenomicRanges)
})
ah <- AnnotationHub()
```

Download all peaks from Mus musculus strain Bruce4 ES-Bruce4

```{r}
p300 <- rtracklayer::import("/Users/catherinerohner/ETH/Bioinformatics/ENCFF653YVR.bed", format="narrowPeak")
H3K4me3 <- rtracklayer::import("/Users/catherinerohner/ETH/Bioinformatics/ENCFF247GVM.bed", format="narrowPeak")
H3K4me1 <- rtracklayer::import("/Users/catherinerohner/ETH/Bioinformatics/ENCFF333IJH.bed", format="narrowPeak")
H3K27ac <- rtracklayer::import("/Users/catherinerohner/ETH/Bioinformatics/ENCFF360VIS.bed", format="narrowPeak")
H3K27me3 <- rtracklayer::import("/Users/catherinerohner/ETH/Bioinformatics/ENCFF008XKX.bed", format="narrowPeak")
```

Create a Function to calculate the proportion

```{r}
calculate_overlap_proportion <- function(p300, histone_mark) {
  overlaps <- findOverlaps(p300, histone_mark)
  proportion <- length(unique(queryHits(overlaps))) / length(p300)
  return(proportion)
}
```

Call out Function for all Histone Marks

```{r}
proportion_H3K4me3 <- calculate_overlap_proportion(p300, H3K4me3)
proportion_H3K4me1 <- calculate_overlap_proportion(p300, H3K4me1)
proportion_H3K27ac <- calculate_overlap_proportion(p300, H3K27ac)
proportion_H3K27me3 <- calculate_overlap_proportion(p300, H3K27me3)

print(proportion_H3K4me3)
print(proportion_H3K4me1)
print(proportion_H3K27ac)
print(proportion_H3K27me3)
```
Discussion: the large overlap in the H3K27ac histone modification shows a strong association to the p300 Transcription Factor. As H3K27ac is a mark for enhancers and promotors, this indicates that p300 has a role in transcriptional activation.


